---
output: github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "##",
  fig.path = "man/images/"
)
```
```{r echo=FALSE, results="hide", message=FALSE}
library("badger")
```

# quanteda.tidy

<!-- badges: start -->
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/quanteda.tidy)](https://cran.r-project.org/package=quanteda.tidy)
`r badge_devel("quanteda/quanteda.tidy", "royalblue")`
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![R build status](https://github.com/quanteda/quanteda.tidy/workflows/R-CMD-check/badge.svg)](https://github.com/quanteda/quanteda.tidy/actions)
[![Coverage Status](https://img.shields.io/codecov/c/github/quanteda/quanteda.tidy/master.svg)](https://codecov.io/github/quanteda/quanteda.tidy?branch=master)
<!-- badges: end -->


## About

**quanteda.tidy** extends the **quanteda** package with functionality from the "tidyverse", especially **dplyr**.  
Note that this is not the same as **tidytext**, which stretches tokens into data.frames.  Instead, tidy functions operate only on document variables, but extends these functions (from **dplyr**) to work on **quanteda** objects as if they were tibbles or data.frames.

## Installation

You can install **quanteda.tidy** from GitHub with:

```{r eval = FALSE}
devtools::install_github("quanteda/quanteda.tidy")
```


## Examples

Adding a document variable for full president name:
```{r}
library("quanteda.tidy", warn.conflicts = FALSE)

data_corpus_inaugural %>%
  transmute(fullname = paste(FirstName, President, sep = ", ")) %>%
  summary(n = 5)

data_corpus_inaugural %>%
  mutate(fullname = paste(FirstName, President, sep = ", ")) %>%
  summary(n = 5)
```

Filtering documents based on years:
```{r}
data_corpus_inaugural %>%
  filter(President == "Roosevelt") %>%
  summary()
```

Renaming document variables:
```{r}
data_corpus_inaugural %>%
  rename(LastName = President) %>%
  select(FirstName, LastName) %>%
  summary(n = 5)
```

Glimpse (from **tibble**):
```{r}
glimpse(data_corpus_inaugural)
```

Slice operations:
```{r}
slice(data_corpus_inaugural, 1:3)

slice_head(data_corpus_inaugural, prop = .10)
slice_tail(data_corpus_inaugural, 3)

set.seed(42)
slice_sample(data_corpus_inaugural, prop = .50)
```
